---
title: "Dashboard Setup Example"
output: html_document
date: "2024-11-28"
---

# 1. Setup

This section handles the installation and loading of necessary packages, as well as loading and preparing the data.

We start by loading all the necessary tools and libraries required for the dashboard functionality. We list all the packages we need and include a handy function to check if any are missing. If it finds any missing packages, it installs them automatically. After ensuring all packages are available, it loads them into the R session.

Next, the code specifies the location of the data file on the computer. It verifies whether the file exists to prevent errors during execution. Once confirmed, it reads the data from the Excel file into R. The data is then cleaned by removing any incomplete entries and ensuring that the year and value columns are in the correct format for analysis and plotting.


```{r}

# List of required packages
required_packages <- c(
  "shiny", 
  "readxl", 
  "dplyr", 
  "ggplot2", 
  "plotly", 
  "shinycssloaders", 
  "DT", 
  "countrycode", 
  "RColorBrewer"
)

# Function to install any missing packages
install_if_missing <- function(packages) {
  missing_packages <- packages[!(packages %in% installed.packages()[, "Package"])]
  if (length(missing_packages)) {
    install.packages(missing_packages, dependencies = TRUE)
  }
}

# Install missing packages
install_if_missing(required_packages)

# Load all packages into the R session
lapply(required_packages, library, character.only = TRUE)

# Path to the data file on your computer
file_path <- "C:/Users/Michael Beshay/Downloads/combined_health_nutrition_population_data_with_categories.xlsx"

# Check if the data file exists
if (!file.exists(file_path)) {
  stop("Data file not found. Please check the file path.")
}

# Read the Excel data into R
df <- read_excel(file_path)

# Clean and prepare the data
df_cleaned <- df %>%
  # Remove rows with missing important information
  filter(
    !is.na(`Country Name`),
    !is.na(`Series Name`),
    !is.na(Year),
    !is.na(Value),
    !is.na(wealth_quintiles),
    !is.na(Category)
  ) %>%
  # Convert Year to integers and Value to numeric
  mutate(
    Year = as.integer(Year),
    Value = as.numeric(Value)
  )

```



# 2. User Interface (UI)

The dashboard layout is divided into two main parts: a sidebar and a main panel. The sidebar contains various input controls that allow users to filter the data. Users can select a country from a dropdown menu, choose one or more categories via checkboxes, pick a specific metric using radio buttons, and define the range of years with a slider. They can also select the type of plot they prefer (Line, Bar, or Scatter) and decide whether to add an average line to the plot. A reset button is provided to revert all selections to their default states.

The main panel hosts two tabs. The first tab, "Visualization," displays the interactive plot along with a loading spinner that appears while the plot is being generated. Below the plot, there's a note indicating the data source. The second tab, "Statistics," shows a table of the filtered data and includes a button for downloading the data as a CSV file.


```{r setup, include=FALSE}

# Define the User Interface (UI) of the dashboard
ui <- fluidPage(
  
  # Add custom styles to make the dashboard look nicer
  tags$head(
    tags$style(HTML("
      /* Style the main title */
      .custom-title {
        font-size: 24px;
        font-weight: bold;
        text-align: center;
        margin-bottom: 20px;
      }
      
      /* Set the height of the plot area */
      #plotContainer {
        height: 400px; /* Smaller height for the plot */
      }
      
      /* Add some padding to the main area */
      .main-panel {
        padding: 10px;
      }
      
      /* Make the text in the dashboard a bit larger */
      body, label, .control-label, .shiny-input-container {
        font-size: 14px;
      }
      
      /* Make the tab titles bigger */
      .nav-tabs > li > a {
        font-size: 16px;
      }
      
      /* Style the notifications */
      .shiny-notification {
        font-size: 14px;
      }
      
      /* Style the source text below the plot */
      .source-text {
        font-size: 12px;
        color: grey;
        text-align: right;
        margin-top: 10px;
      }
    "))
  ),
  
  # Set the main title of the dashboard
  titlePanel(
    div(
      "Metrics by Wealth Quintiles",
      class = "custom-title"
    )
  ),
  
  # Create a layout with a sidebar and main area
  sidebarLayout(
    sidebarPanel(
      width = 3, # Make the sidebar take up 3 columns
      
      # Dropdown menu to select a country
      selectInput(
        inputId = "country",
        label = "Select a Country:",
        choices = sort(unique(df_cleaned$`Country Name`)),
        selected = sort(unique(df_cleaned$`Country Name`))[1]
      ),
      
      # Checkboxes to select one or more categories
      checkboxGroupInput(
        inputId = "category",
        label = "Select Categories:",
        choices = sort(unique(df_cleaned$Category)),
        selected = sort(unique(df_cleaned$Category))[1] # Choose the first category by default
      ),
      
      # Radio buttons to select a single metric
      radioButtons(
        inputId = "series",
        label = "Metrics:",
        choices = sort(unique(df_cleaned$`Series Name`)), # List all available metrics
        selected = sort(unique(df_cleaned$`Series Name`))[1] # Choose the first metric by default
      ),
      
      # Slider to select the range of years
      sliderInput(
        inputId = "year_range",
        label = "Select Year Range:",
        min = min(df_cleaned$Year, na.rm = TRUE),
        max = max(df_cleaned$Year, na.rm = TRUE),
        value = c(min(df_cleaned$Year, na.rm = TRUE), max(df_cleaned$Year, na.rm = TRUE)),
        step = 1,
        sep = "" # Remove thousand separators
      ),
      
      # Radio buttons to choose the type of plot
      radioButtons(
        inputId = "plot_type",
        label = "Select Plot Type:",
        choices = c("Line Plot", "Bar Plot", "Scatter Plot"),
        selected = "Line Plot",
        inline = TRUE
      ),
      
      # Checkbox to add an average line to the plot
      checkboxInput(
        inputId = "add_avg_line",
        label = "Add Black Dotted Average Line",
        value = FALSE # Not checked by default
      ),
      
      # Button to reset all filters to their default values
      actionButton(
        inputId = "reset_filters",
        label = "Reset Filters",
        icon = icon("redo"),
        class = "btn-primary"
      )
    ),
    
    mainPanel(
      class = "main-panel",
      tabsetPanel(
        # First tab for visualization
        tabPanel(
          "Visualization",
          br(), # Add a line break for spacing
          # Container for the plot with a set height
          div(
            id = "plotContainer",
            plotlyOutput("plot", height = "100%") %>% withSpinner(color = "#0dc5c1") # Show a loading spinner
          ),
          # Text to show the data source below the plot
          div(
            "Source: World Bank - Health, Nutrition, and Population Statistics",
            class = "source-text"
          )
        ),
        
        # Second tab for statistics
        tabPanel(
          "Statistics",
          br(), # Add a line break for spacing
          # Button to download the data as a CSV file
          downloadButton("downloadData", "Download CSV", class = "btn-success"),
          br(), br(), # Add two line breaks for spacing
          # Table to show the statistics with a loading spinner
          DTOutput("statsTable") %>% withSpinner(color = "#0dc5c1")
        )
      )
    )
  )
)


```



# 3. Server Logic

This section contains the backend functionality, which includes data filtering, plot creation, and handling user interactions.

Here's what each part does:

Metric Selection Update: When users select different categories, this part updates the list of available metrics accordingly. If no metrics are available for the chosen categories, it notifies the user with an error message.

Reset Functionality: If users click the "Reset Filters" button, this section restores all input controls to their default settings.

Data Filtering: Based on the user's selections for country, categories, metric, and year range, this reactive function filters the dataset to include only the relevant data.

Plot Preparation: After filtering, this part calculates the average value for each year. This average is used later to add a dashed average line to the plot if the user chooses to.

Plot Creation: Depending on the user's choice of plot type (Line, Bar, or Scatter), this section builds the appropriate plot using ggplot2. It assigns specific colors and shapes to different wealth quintiles for clarity. 

Statistics Table: This part displays the filtered data in a neat, interactive table. Users can search, sort, and navigate through the data. Additionally, there's an option to download the table as a CSV file for offline analysis - if not desired we can exclude it. 


```{r}

# Define the server part of the dashboard
server <- function(input, output, session) {
  
  # Update the list of metrics based on selected categories
  observeEvent(input$category, {
    if (length(input$category) > 0) { # If any category is selected
      metrics_filtered <- df_cleaned %>%
        filter(Category %in% input$category) %>% # Keep only selected categories
        pull(`Series Name`) %>% 
        unique() %>% # Get unique metrics
        sort()
      
      if (length(metrics_filtered) > 0) { # If there are metrics available
        updateRadioButtons(session, "series", choices = metrics_filtered, selected = metrics_filtered[1])
      } else {
        # If no metrics are available, clear the selection and show a message
        updateRadioButtons(session, "series", choices = character(0), selected = NULL)
        showNotification("No Metrics available for the selected Categories.", type = "error")
      }
    } else {
      # If no categories are selected, show all metrics
      all_metrics <- sort(unique(df_cleaned$`Series Name`))
      if (length(all_metrics) > 0) {
        updateRadioButtons(session, "series", choices = all_metrics, selected = all_metrics[1])
      } else {
        # If no metrics are available at all, clear the selection and show a message
        updateRadioButtons(session, "series", choices = character(0), selected = NULL)
        showNotification("No Metrics available.", type = "error")
      }
    }
  }, ignoreNULL = FALSE) # Run this when the dashboard starts
  
  # Reset all filters to their default values when the button is clicked
  observeEvent(input$reset_filters, {
    updateSelectInput(session, "country", selected = sort(unique(df_cleaned$`Country Name`))[1])
    updateCheckboxGroupInput(session, "category", selected = sort(unique(df_cleaned$Category))[1]) # Reset to first category
    updateRadioButtons(session, "series", selected = sort(unique(df_cleaned$`Series Name`))[1]) # Reset to first metric
    updateSliderInput(session, "year_range",
                      value = c(min(df_cleaned$Year, na.rm = TRUE), max(df_cleaned$Year, na.rm = TRUE)))
    updateRadioButtons(session, "plot_type", selected = "Line Plot")
    updateCheckboxInput(session, "add_avg_line", value = FALSE)
  })
  
  # Function to filter the data based on user inputs
  filtered_data_sidebar <- reactive({
    data <- df_cleaned
    
    # Keep only the selected country
    if (!is.null(input$country) && input$country != "") {
      data <- data %>% filter(`Country Name` == input$country)
    }
    
    # Keep only the selected categories
    if (length(input$category) > 0) {
      data <- data %>% filter(Category %in% input$category)
    }
    
    # Keep only the selected metric
    if (!is.null(input$series) && input$series != "") {
      data <- data %>% filter(`Series Name` == input$series)
    }
    
    # Keep only the data within the selected year range
    data <- data %>%
      filter(Year >= input$year_range[1], Year <= input$year_range[2])
    
    return(data)
  })
  
  # Prepare the data for plotting
  plot_data_prepared <- reactive({
    data <- filtered_data_sidebar()
    
    # If there's no data after filtering, return NULL
    if (nrow(data) == 0) {
      return(NULL)
    }
    
    # Calculate the average value for each year
    avg_data <- data %>%
      group_by(Year) %>%
      summarize(
        Avg_Value = mean(Value, na.rm = TRUE),
        .groups = 'drop'
      )
    
    return(list(main = data, average = avg_data))
  })
  
  # Create the interactive plot
  output$plot <- renderPlotly({
    plot_data <- plot_data_prepared()
    
    # If there's no data, show a warning and stop
    if (is.null(plot_data)) {
      showNotification("No data available for the selected filters. Please adjust your selections.", type = "warning")
      return(NULL)
    }
    
    main_data <- plot_data$main
    avg_data <- plot_data$average
    
    # Get all unique wealth quintiles
    quintiles <- sort(unique(main_data$wealth_quintiles))
    
    # Choose colors for each quintile
    num_quintiles <- length(quintiles)
    if (num_quintiles <= 9) { # Brewer has up to 9 colors
      quintile_colors <- brewer.pal(num_quintiles, "Set1")
    } else {
      # If more than 9, create a larger palette
      quintile_colors <- colorRampPalette(brewer.pal(9, "Set1"))(num_quintiles)
    }
    
    # Assign colors to quintiles and the average line
    color_values <- c(quintile_colors, "Average Line" = "black")
    names(color_values) <- c(quintiles, "Average Line")
    
    # Start building the plot
    p <- ggplot(main_data, aes(x = Year, y = Value, color = wealth_quintiles, shape = wealth_quintiles, 
                               text = paste0("Year: ", Year, "<br>Value: ", round(Value, 2)))) +
      labs(
        x = "Year",
        y = "Value",
        color = "Wealth Quintile",
        shape = "Wealth Quintile",
        title = paste(main_data$`Country Name`[1], "-", input$series) # Title as "Country - Variable"
      ) +
      theme_minimal() +
      scale_color_manual(values = color_values) + # Use the defined colors
      scale_shape_manual(values = c(16, 17, 15, 3, 7, 8, 9, 10, 11)) + # Different shapes for quintiles
      theme(
        legend.position = "bottom", # Place legend at the bottom
        plot.title = element_text(size = 16, hjust = 0.5, vjust = 1, face = "bold") # Style the title
      )
    
    # Add the type of plot chosen by the user
    if (input$plot_type == "Line Plot") {
      p <- p + geom_line(aes(group = wealth_quintiles), size = 1) # Draw lines
      p <- p + geom_point(size = 3) # Add points on the lines
    } else if (input$plot_type == "Bar Plot") {
      p <- p + geom_bar(stat = "identity", position = position_dodge()) # Create bar plot
    } else if (input$plot_type == "Scatter Plot") {
      p <- p + geom_point(size = 3, alpha = 0.7) # Create scatter plot
    }
    
    # Add the average line if the user checked the box
    if (input$add_avg_line) {
      p <- p + 
        geom_line(data = avg_data, aes(x = Year, y = Avg_Value, color = "Average Line"), 
                  linetype = "dashed", size = 1) + # Draw dashed average line
        geom_point(data = avg_data, aes(x = Year, y = Avg_Value, color = "Average Line"), 
                   size = 3, shape = 8) # Add points for average line
    }
    
    # Make the plot interactive with Plotly
    ggplotly(p, tooltip = "text") %>%
      layout(
        title = list(
          text = paste(main_data$`Country Name`[1], "-", input$series),
          font = list(size = 16)
        ),
        legend = list(
          orientation = "h", # Horizontal legend
          x = 0.5, # Center horizontally
          y = -0.3, # Position below the plot
          xanchor = "center",
          yanchor = "top"
        )
      ) %>%
      config(displayModeBar = FALSE) # Hide the toolbar for a cleaner look
  })
  
  # Show the statistics table
  output$statsTable <- renderDT({
    data <- filtered_data_sidebar()
    
    # If there's no data, don't show the table
    if (nrow(data) == 0) {
      return(NULL)
    }
    
    # Create a table with selected columns
    datatable(
      data %>%
        select(`Country Name`, Category, `Series Name`, Year, Value, wealth_quintiles),
      options = list(pageLength = 10, scrollX = TRUE), # Show 10 rows per page and allow horizontal scrolling
      rownames = FALSE, # Don't show row numbers
      filter = "top" # Add filters at the top of the table
    )
  })
  
  # Allow users to download the data as a CSV file
  output$downloadData <- downloadHandler(
    filename = function() {
      paste("health_nutrition_population_data_", Sys.Date(), ".csv", sep = "") # Name the file with the current date
    },
    content = function(file) {
      write.csv(filtered_data_sidebar(), file, row.names = FALSE) # Write the data to the file
    }
  )
}

```



# 4. Running the dashboard


```{r}
# Start the Shiny app
shinyApp(ui = ui, server = server)

```


